HyperLogLog: Exponentially Bad in Adversarial Settings

### 2021-04-30 docs/HLL_exp_bad/main.tex
work on intro

\section{Introduction}
\todo{more here introducing big data and why we want statistical estimators on them; data streams; cannot store, or large data sets, want summary statistics. Such things are supported by GBQ, Apache Spark, Redis. Cardinality as a key metric; HLL as the algorithm of choice. ''analytics engines for big data processing.''}
Naive approaches to computing the cardinality of large data sets, such as sorting the elements or simply maintaining the set of unique elements seen, are impractical at scale and in streaming settings. The HyperLogLog algorithm (HLL henceforth)~\cite{hll} is today the most widely used cardinality estimator. It is increasingly used in settings where adversaries may have  incentives to manipulate the estimate made by the algorithm. For example:
\begin{itemize}
\item Google and counting stuff.\todo{add more detail here -- their blogs, their paper, the use in Google Big Query.} 
    \item Cardinality estimation provides a significantly faster way for Facebook to compute statistics on their users. They use HLL to find out how many distinct people visited their website in the past week~\cite{fbhll}.
    \item HLL has been proposed as a solution to detecting Denial-of-Service attacks~\cite{portscanhll} (this paper actually uses a variation on HLL called \emph{sliding HLL}~\cite{slidinghll} which estimates the cardinality over a moving time window). Specifically, HLL is used to count how many different ports are used as destination port over all packets received, as an attempt to identify when a port scan attack is underway.
    \item HLL has been implemented in network soft-switches to approximate the number of distinct packets in traffic flows and overcome the switch resource limits~\cite{flexswitch}. The number of unique flows traversing a switch can then be used to provide a better congestion control protocol.
\end{itemize}

Regardless of the sensitivity of the above examples, none of the above-cited sources are clear about their threat model, especially for the HLL component. They are all potentially vulnerable to the attacks we present in this paper.

Recently, Reviriego and Ting~\cite{hllvuln} initiated work on the performance of HLL in adversarial settings. They showed that in very specific attack settings, the cardinality estimate made by HLL can be modified through input selection. In this paper, we present a complete analysis of HLL in adversarial settings, considering more realistic attack scenarios and giving more powerful attacks than~\cite{hllvuln}. In particular, we show that with only modest knowledge of the HLL internals and a moderate amount of computation, an adversary with sufficient flexibility in the choice of inputs can make the HLL cardinality estimate exponentially smaller than the true cardinality (e.g.\ constant instead of ${\mathcal{O}}(2^t)$ where $t$ is the number of bits of flexibility in the adversary's input). We demonstrate this theoretically and experimentally for the ``classic'' version of HLL. We also study a specific ``modern'' version of HLL which adopts recommendation from~\cite{hllnew} and show how variants of our attacks apply there too. This modern version of HLL is interesting because it is the one used in Redis, a popular key-value database technology.

To complement our attack results, we provide a formal security analysis of HLL. We show that in the ``shadow device'' attack scenario of~\cite{hllvuln}, no security is possible even if HLL is secretly keyed. On the other hand, we also show that in the most powerful attack setting introduced here, where the attacker has full access to the internals of the HLL algorithm (except for a secret key!), it is possible to provably secure HLL against cardinality manipulation by the simple expedient of replacing its internal hash function by a keyed, variable-input-length pseudo-random function (VIL-PRF). We discuss and benchmark a suitable choice for the required VIL-PRF.

\todo[inline]{make the model a contribution... ``To achieve this... introduce simulation-based security model for cardinality estimators'', etc}
\todo[inline]{Simplicity of attacks as a strength; only complexity comes in S1 attack which we don't consider that interesting!}
\todo[inline]{No e2e attack on a specific system; rather paper is written in general form since the attacks apply broadly.}
\todo[inline]{Highlight the analysis of OSDI paper~\cite{flexswitch} as an example of the potential of the attacks.}

\subsection{Paper Organisation}
The rest of the paper is organised as follows. Section~\ref{sec:related} immediately below discusses related work. Section~\ref{sec:overview} gives an overview of HLL. Section~\ref{sec:attacks} presents our different adversarial models. In Section~\ref{sec:manip}, we describe attacks on HLL and evaluate their impacts under the different adversarial models. Section~\ref{sec:redis} considers the security of the more modern version of HLL due to Ertl~\cite{hllnew} that is used in Redis. Our experimental results are discussed and related back to specific use cases in Section~\ref{sec:exp}. Section~\ref{sec:formal} provides our formal security analysis of HLL. Section~\ref{sec:conclusions} provides our conclusions and some ideas for future work.

@@

So to roughly estimate the cardinality of a stream, we need only to store a representation of $k$, the length of the largest observed string of leading zero-bits for items in the stream. This provides a very compact mechanism for estimating a stream's cardinality -- if the true cardinality is $N$, then just $\mathcal{O}(\log\log(N))$ bits of storage are needed to compute the cardinality estimate.

This simple approach suffers from large variance. In order to reduce the variance, HLL uses many estimators in parallel instead of one and averages the results. Each estimate is stored in its own register that we call a bucket. We refer to the collection of buckets and their contents as the HLL \emph{sketch}. To map a value $x$ to one of $m=2^n$ buckets, \cite{loglog} suggests using the first $n$ bits of $h(x)$ as a bucket index, and then computing the longest sequence of leading zero bits on the remaining $\ell$ bits of $h(x)$. The output length of $h(\cdot)$ is thus $n+\ell$. This sum is fixed to 32 in the original HLL paper~\cite{hll}. Estimates from all buckets are averaged using the harmonic mean and scaled by a constant $\alpha_m$, where $\alpha_m$ is empirically computed to correct a systematic multiplicative bias.

In the original HLL proposal~\cite{hll}, relatively small and large cardinalities are handled separately to correct for systematic errors. When the raw cardinality estimate is below $\frac{5}{2}m$, the Hit Counting algorithm of Whang \emph{et al.}~\cite{hitcounting} is used to produce the final cardinality estimate. This estimator uses the number of empty buckets (whose register value is 0) instead of the number of leading zeros in the stream. A similar correction is applied for large cardinalities in order to take into account the likelihood of hash collisions. In addition, the cardinality estimate is usually rounded to a whole number before being output. The original HLL algorithm from~\cite{hll} is shown in Fig.~\ref{fig:hll}. This algorithm is able to estimate cardinalities greater than $10^9$ with a typical standard error of 4\%, using only 1.5 kB of memory. Crucially, the proof of this~\cite[Theorem 1]{hll} makes the assumption that the inputs are drawn independently and uniformly at random from the input domain.\footnote{Technically, the results of~\cite{hll} are proven for a distribution obtained by making arbitrary replications and permutations after sampling independently and uniformly, to simulate input taken from an \emph{ideal multiset}. But they also hold without replication and permutation because HLL's internal state is invariant under repeated inputs.}

@@

\subsection{Attacker's Objective, Abstractly}
Throughout, we assume the adversary has the ability to vary the contents of free fields in the strings which will be inserted into the stream of values. For example it might manipulate IPID and other header fields in the context of IP packets. We model this in a general way by assuming the adversary has access to a large set ${\mathcal{X}}$ of potential inputs from which it can select items to insert into the HLL sketch. Our adversary's objective, then, is to insert as many items from ${\mathcal{X}}$ as possible whilst keeping the HLL cardinality estimate as low as possible.

@@

In this setting, where the target HLL sketch is initially non-empty, we see that an adversary who has sufficient flexibility in its inputs (as represented by the size of the set ${\mathcal{X}}$) can expect to insert many values into the HLL sketch without increasing the cardinality estimate (or whilst increasing the cardinality estimate slightly in the case where the HLL sketch has some empty buckets in S2). The result is largely independent of the HLL parameters. The more full is the HLL sketch, the larger is $1-H_M^{-1}$ and the easier it is to insert new items without increasing the cardinality estimate.

@@

Since the target sketch has an unknown status for the adversary, our previous strategy for S1 of keeping many buckets empty no longer works. One can attempt to use the shadow device to identify items that would result in small entries $M[j]$ (for unknown indices $j$) since such items are unlikely to increase the cardinality estimate of the target device by much.  This can be done by first filling the shadow device with sufficiently many values to escape from the ``small range correction'' phase. Then one could enter candidate values $x$ into the shadow device and check for cardinality increases. However this runs into the rounding issue described in Section~\ref{sec:scenarios-discussion}. Still, one could retain those items which do lead to a detectable change in cardinality and reject the others (and also estimate the value of $M[j]$ for the unknown bucket $j$ from the magnitude of the change, retaining only those for which $M[j]$ is small). But here the lack of a cardinality change could result either from rounding issues or because insertion of candidate value $x$ into the shadow device really did lead to no increase in cardinality. This indicates that  ``good'' items would be rejected unnecessarily. This could be addressed by using multiple resets of the shadow HLL sketch, filling it with random items after each reset, and performing multiple insertions per candidate item $x$ to try to estimate the true value of $M[j]$ for each item $x$ (since after at least some of the resets followed by random insertions, we could expect $M[j]$ to increase when $x$ is inserted, and if $M[j]$ is small enough relative to the cardinality estimate, the change could be detected). The details of such an attack require further analysis which we do not pursue here, since we consider S1 to be the least realistic attack scenario.

%Finally, we briefly consider scenario S1 with an initially non-empty target HLL sketch. Since the target sketch has an unknown status for the adversary, our previous strategy for S1 of keeping many buckets empty does not work. One can attempt to use the shadow device to identify items that would result in small entries $M[j]$ (for unknown indices $j$) since such items are unlikely to increase the cardinality estimate of the target device by much.  This requires first filling the shadow device with sufficiently many values to escape from the ``small range correction'' phase. Then one could enter random values $x$ into the shadow device and check for cardinality increases. However this runs into the rounding issue described in Section~\ref{sec:scenarios-discussion}. Still, one could retain those items which do lead to a detectable change in cardinality and reject the others (and also estimate the value of $M[j]$ for the unknown bucket $j$ from the magnitude of the change, retaining only those for which $M[j]$ is small). But here the lack of a cardinality change could result either from rounding issues or because insertion of $x$ into the shadow device really did lead to no increase in cardinality. This indicates that  ``good'' items would be rejected unnecessarily. This could be addressed by using multiple resets, filling with random items, and then multiple insertions per item to try to estimate the true value of $M[j]$ for each item. The details of such an attack require further analysis which we do not pursue here, since we consider S1 to be the least realistic attack scenario.

@@

In scenarios S2 and S4 with an initially non-empty HLL sketch, we assume as for our earlier analysis of the standard HLL algorithm that all buckets have already been hit (this will be the case with high probability as soon as $m \log m$ items have been honestly inserted). Hence $C_0 = 0$ and $\sigma$ is evaluated at 0 in (\ref{eqn:ertl}), where its value is 0. This means that the estimate $\hat{\lambda}$ is computed using a denominator of the form:
\[
\sum_{k=1}^{\ell}C_k 2^{-k} + m \tau(1-C_{\ell+1}/m)2^{-\ell}.
\]
Now we can proceed exactly as in Sections~\ref{sec:nonemptyHLLS2} and~~\ref{sec:nonemptyHLLS4}, inserting values $x$ whose hashes have rightmost $\ell$ bits beginning with a ``1'' bit (and with a more relaxed condition in S4). These insertions do not alter the values of the $C_k$, and hence $\hat{\lambda}$ is unaffected. In S2, we expect to insert half the items in ${\mathcal{X}}$ without increasing the cardinality estimate $\hat{\lambda}$ at all. In S4, the analysis is the same as in Section~\ref{sec:nonemptyHLLS4}.

@@

We first begin by explaining the high-level idea of our approach. We allow an adversary ${\mathcal{A}}$ to interact with an HLL via various oracles that capture its capabilities in each of the three scenarios S2, S3 and S4: it may have an \texttt{insert($\cdot$)} oracle allowing it to add values to the HLL sketch, a \texttt{read} oracle that returns the HLL cardinality estimate, and a \texttt{corrupt} that returns all the HLL bucket contents (but not any secret key $K$); in S4 it has all three oracles. In addition, in all 3 scenarios, ${\mathcal{A}}$ has access to an \texttt{init($\cdot$)} oracle by which it sets the HLL parameters $n$, $\ell$, $\alpha_n$ used to initialise the HLL sketch. We assume ${\mathcal{A}}$ always makes exactly one \texttt{init($\cdot$)} query and that it is its first oracle call.

@@

The four oracles that  ${\mathcal{A}}$ can interact with behave as follows in the real world:
\begin{itemize}
\item \texttt{init($\cdot$)}: this oracle receives as input the HLL-$F$ parameters ($n$, $\ell$, $\alpha_n$) and performs initialisation of the HLL-$F$ sketch, including generating the array of buckets $M[j]$ and initialising all the entries to $0$, and selecting a key $K \leftarrow_{\$} \{0,1\}^k$ for keyed function $F: \{0,1\}^k \times {{\mathcal{D}}} \rightarrow  \{0,1\}^{n+\ell}$. This oracle has no output. It is available in all of S2, S3, S4. We assume it is only called once as ${\mathcal{A}}$'s first oracle query.
\item \texttt{insert($\cdot$)}: on input $x \in {{\mathcal{D}}}$, the oracle implements HLL-$F$ by setting $y = F_K(x)$ and following the procedure in Figure~\ref{fig:hll} to insert value $x$ into the HLL sketch (that is, it uses the first $m$ bits of $y$ to determine a bucket index $j$ and the next $\ell$ bits of $y$ to determine a bit-string $w$; then $\rho(w)$ is used to update the bucket at index $j$ via $M[j]:= \max(M[j],\rho(w))$). This oracle has no output. It is available in all of S2, S3, S4. Without loss of generality (since HLL and variants of it we consider here have a deterministic insertion procedure), we assume all queries $x$ made by ${\mathcal{A}}$ to \texttt{insert($\cdot$)} are distinct.
\item \texttt{read}: this oracle has no input and returns the HLL-$F$ cardinality estimate as per Figure~\ref{fig:hll}, i.e.\ it computes a raw cardinality estimate via $E := \alpha_mm^2\cdot(\sum_{j=1}^m2^{-M[j]})^{-1}$ and then performs corrections for small and large values. We allow multiple queries to this oracle and make it available to ${\mathcal{A}}$ in S3 and S4, but not S2.
\item \texttt{corrupt}: this oracle has no input and returns the vector $(M[j])_{j=1}^{m}$ of HLL bucket contents (but not any secret key $K$). This oracle is available only in S4. We allow the adversary to query \texttt{corrupt} as often as it wishes during its attack. Our description of S4 in Section~\ref{sec:attacks} limited the adversary to at most one such query; here we allow multiple queries because doing so does not increase the difficulty of achieving security.  
\end{itemize}

@@

An HLL can leak information about values $x$ that were previously inserted. Consider the following attack from~\cite{cardestprivacy}: to test whether a value $x$ was previously inserted into the HLL by an honest user (say), the adversary makes the sequence of queries: \texttt{read}, \texttt{insert($x$)}, \texttt{read}. If the two cardinality estimates are different, the adversary can conclude definitively that $x$ was \emph{not} previously inserted into the HLL; if they are the same, it can estimate (with some error) that $x$ was previously inserted.\footnote{Errors can arise in two ways when $x$ was not previously inserted: either inserting $x$ changes a bucket value, but the change is masked by rounding when the cardinality estimate is computed, or inserting $x$ does lead to a change in a bucket value because $\rho(w)$ computed from $x$ is too small.} This attack still works even if HLL is implemented using a PRF and is unavoidable in view of the oracle queries (and in real settings, the API) available to the adversary. So such privacy leaks are an inherent feature of HLL if a rich enough API is provided. Note that this attack does not contradict our main result, which concerns prevention of cardinality manipulation rather than confidentiality.

%Still, simulation-based security notions are often used to model confidentiality properties, so it is tempting to think that our security definition might say something about confidentiality of the HLL inputs $x$. Our existing model does not, since ${\mathcal{A}}$ chooses all the inputs $x$ and so ``knows'' them already. However, it is easily modified to provide a confidentiality notion. For example, we can consider the insertion of honestly generated values $x$ that are drawn from some probability distribution $p$ on $\mathcal{D}$ in a first phase, with ${\mathcal{A}}$ then having access to only \texttt{read} and \texttt{corrupt} oracles in a second phase (we have to remove the \texttt{insert($\cdot$)} oracle in view of the general attack from~\cite{cardestprivacy}).

@@

An S4 adversary $\mathcal{A}$ can perform a series of pairs of \texttt{insert($x^i$)} and \texttt{corrupt} queries on chosen values $x^0, x^1, \ldots, x^{t-1}$. With high probability (assuming $t$ is not too large), each \texttt{insert} query hits a fresh bucket in the HLL sketch. Then comparing the results of the \texttt{corrupt} queries either side of the \texttt{insert($x^i$)} query reveals to $\mathcal{A}$ which bucket was hit by that  \texttt{insert} query. In turn, this leaks partial information about the corresponding UHF output $p_{x^i}(K)$, specifically, its $n$ least significant bits (since those bits are used to generate the bucket index for insertion). We now show how to exploit this partial information on $p_{x^i}(K)$ to recover $K$.

@@

\todo{Sim based of indepednent interest, arguably more powerful than game-based notions with fixed winning predicates.}
%It is also an open problem to close the performance gap that exists between fast but non-cryptographic hashes used in HLL implementations today and the slower but cryptographically strong VIL-PRF that is needed to obtain security.

>>>>

\section{Introduction}
Data scientists today make use of systems like Google Big Query, Apache Spark, Presto and Redis to handle large data sets. Such systems include analytics engines capable of performing powerful operations on the data. Particularly useful to data scientists are summary statistics, with a key statistic being the number of distinct data items in a data set, i.e.\ the data set's cardinality. In many use cases, obtaining an approximate cardinality is sufficient. Over the last decade, the HyperLogLog (HLL) cardinality estimator~\cite{hll} has emerged as the algorithm of choice for this task, due to its excellent speed, low memory requirements and high proven accuracy. It is also suitable for use in applications where it is not possible (or desirable) to store an entire data set, but instead where the data set is presented in a streaming fashion. HLL is used by all of the above-mentioned systems; section~\ref{sec:overview} contains a complete specification of HLL.

However, HLL is increasingly used in settings where adversaries may have incentives to manipulate the cardinality estimates made by the algorithm, i.e.\ HLL is being used in adversarial settings. For example:
\begin{itemize}
    \item Cardinality estimation provides a significantly faster way for Facebook to compute statistics on their users. They use HLL to find out how many distinct people visited their website in the past week~\cite{fbhll}.
    \item Google uses a variant of HLL described in~\cite{hllpractice} in its PowerDrill system (first described in~\cite{HallBBGN12}) and possibly in a variety of other internal systems reported in~\cite{hllpractice}. For example, cardinality estimation is used to to determine the number of distinct search queries on \url{google.com} over a time period~\cite{hllpractice}. HLL is also used for cardinality estimation via an API in Google Big Query. 
    \item HLL has been implemented in network soft-switches to approximate the number of distinct packets in traffic flows and overcome the switch memory limits~\cite{flexswitch}. The number of unique flows traversing a switch is then used to  inform a congestion control mechanism.
    \item HLL has been proposed as a solution to detecting Denial-of-Service attacks~\cite{portscanhll} (this paper actually uses a variation on HLL called \emph{sliding HLL}~\cite{slidinghll} which estimates the cardinality over a moving time window). Specifically, HLL is used to count how many different destination ports are seen over all packets received, in an attempt to identify when a port scan attack is underway.
\end{itemize}

None of the above-cited examples discuss any kind of threat model where an adversary may try to manipulate the HLL cardinality estimate to its advantage, for example, through input selection. This is despite them all operating in potentially adversarial settings. We infer that the developers of these systems may not be fully aware of the risks they are taking in relying on HLL.

Recently, Reviriego and Ting~\cite{hllvuln} initiated work on the performance of HLL in adversarial settings. They showed that in very specific attack settings, the cardinality estimate made by HLL can be modified through input selection. In this paper, we present a complete analysis of HLL in adversarial settings, considering more realistic attack scenarios and giving more powerful attacks than~\cite{hllvuln}. In particular, we show that with only modest knowledge of the HLL internals and a moderate amount of computation, an adversary with sufficient flexibility in the choice of inputs can make the HLL cardinality estimate exponentially smaller than the true cardinality (e.g.\ constant instead of ${\mathcal{O}}(2^t)$ where $t$ is the number of bits of flexibility in the adversary's input). We demonstrate this theoretically and experimentally for the ``classic'' version of HLL. We also study a specific ``modern'' version of HLL which adopts recommendations from~\cite{hllnew} and show how variants of our attacks apply there too. This modern version of HLL is interesting because it is the one used in Redis, a popular key-value database technology.

We stress that none of our attacks are particularly deep in cryptanalytic terms. Yet HLL is being increasingly widely deployed, and never with protections against our attacks (insofar as we are aware). So despite their technical simplicity, we argue that our attacks are important in establishing the insecurity of HLL in adversarial setting. We present our attacks in an abstract way rather than tying them to any specific setting in order to illustrate their generality. We do give a more detailed exploration of their impact on the data-centre network monitoring system proposed in~\cite{flexswitch}, see Section~\ref{sec:exp}.

To complement our attack results, we provide a formal security analysis of HLL. We show that in the ``shadow device'' attack scenario of~\cite{hllvuln}, no security is possible even if HLL is secretly keyed. On the other hand, we also show that in the most powerful attack setting introduced here, where the attacker has full access to the internals of the HLL algorithm (except for a secret key!), it is possible to provably secure HLL against cardinality manipulation by the simple expedient of replacing its internal hash function $h$ by a keyed, variable-input-length pseudo-random function (VIL-PRF) $F$. We discuss and benchmark a suitable choice for the required VIL-PRF. To support our formal analysis, we introduce a novel simulation-based security model for HLL. We believe our model and approach is of independent interest and will be useful beyond the analysis of cardinality manipulation attacks, for example when studying privacy properties of HLL and other probabilistic data structures.

\subsection{Paper Organisation}
The rest of the paper is organised as follows. Section~\ref{sec:related} immediately below discusses related work. Section~\ref{sec:overview} gives an overview of HLL. Section~\ref{sec:attacks} presents our different adversarial models. In Section~\ref{sec:manip}, we describe attacks on HLL and evaluate their impacts under the different adversarial models. Section~\ref{sec:redis} considers the security of the more modern version of HLL due to Ertl~\cite{hllnew} that is used in Redis. Our experimental results are discussed and related back to a specific application setting in Section~\ref{sec:exp}. Section~\ref{sec:formal} provides our formal security analysis of HLL. Section~\ref{sec:conclusions} provides our conclusions and ideas for future work.

@@

So to roughly estimate the cardinality of a stream, we need only to store a representation of $k$, the length of the largest observed string of leading zero-bits for hashes of items in the stream. This provides a very compact mechanism for estimating a stream's cardinality -- if the true cardinality is $N$, then just $\mathcal{O}(\log\log(N))$ bits of storage are needed to compute the cardinality estimate.

This simple approach suffers from large variance. In order to reduce the variance, HLL uses many estimators in parallel instead of one and averages the results. Each estimate is stored in its own register that we call a bucket. We refer to the collection of buckets and their contents as the HLL \emph{sketch}. To map a value $x$ to one of $m=2^n$ buckets, \cite{loglog} suggests using the first $n$ bits of $h(x)$ as a bucket index, and then computing the longest sequence of leading zero bits on the remaining $\ell$ bits of $h(x)$. The output length of $h(\cdot)$ is thus $n+\ell$. This sum is fixed to 32 in the original HLL paper~\cite{hll}. Estimates from all buckets are averaged using the harmonic mean and scaled by a constant $\alpha_m$, where $\alpha_m$ is empirically computed to correct a systematic multiplicative bias. See Figure~\ref{fig:hll}.

In the original HLL proposal~\cite{hll}, relatively small and large cardinalities are handled separately to correct for systematic errors. When the raw cardinality estimate is below $\frac{5}{2}m$, the Hit Counting algorithm of Whang \emph{et al.}~\cite{hitcounting} is used to produce the final cardinality estimate. This estimator uses the number of empty buckets (whose register value is 0) instead of the number of leading zeros in the stream. A similar correction is applied for large cardinalities in order to take into account the likelihood of hash collisions. In addition, the cardinality estimate is usually rounded to a whole number before being output. The original HLL algorithm from~\cite{hll} is shown in Figure~\ref{fig:hll}. This algorithm is able to estimate cardinalities greater than $10^9$ with a typical standard error of 4\%, using only 1.5 kB of memory. Crucially, the proof of this~\cite[Theorem 1]{hll} makes the assumption that the inputs are drawn independently and uniformly at random from the input domain.\footnote{Technically, the results of~\cite{hll} are proven for a distribution obtained by making arbitrary replications and permutations after sampling independently and uniformly, to simulate input taken from an \emph{ideal multiset}. But they also hold without replication and permutation because HLL's internal state is invariant under repeated inputs.}

@@

\subsection{Attacker's Objective, Abstractly}
Throughout, we assume the adversary has the ability to vary the contents of free fields in the strings which will be inserted into the stream of items. For example it might manipulate IPID and other header fields in the context of IP packets. We model this in a general way by assuming the adversary has access to a large set ${\mathcal{X}}$ of potential inputs from which it can select items to insert into the HLL sketch. Our adversary's objective, then, is to insert as many items from ${\mathcal{X}}$ as possible whilst keeping the HLL cardinality estimate as low as possible.

@@

In this setting, where the target HLL sketch is initially non-empty, we see that an adversary who has sufficient flexibility in its inputs (as represented by the size of the set ${\mathcal{X}}$) can expect to insert many items into the HLL sketch without increasing the cardinality estimate (or whilst increasing the cardinality estimate slightly in the case where the HLL sketch has some empty buckets in S2). The result is largely independent of the HLL parameters. The more full is the HLL sketch, the larger is $1-H_M^{-1}$ and the easier it is to insert new items without increasing the cardinality estimate.

@@

Since the target sketch has an unknown status for the adversary, our previous strategy for S1 of keeping many buckets empty no longer works. One can attempt to use the shadow device to identify items that would result in small entries $M[j]$ (for unknown indices $j$) since such items are unlikely to increase the cardinality estimate of the target device by much.  This can be done by first filling the shadow device with sufficiently many items to escape from the ``small range correction'' phase. Then one could enter candidate items $x$ into the shadow device and check for cardinality increases. However this runs into the rounding issue described in Section~\ref{sec:scenarios-discussion}. Still, one could retain those items which do lead to a detectable change in cardinality and reject the others (and also estimate the value of $M[j]$ for the unknown bucket $j$ from the magnitude of the change, retaining only those for which $M[j]$ is small). But here the lack of a cardinality change could result either from rounding issues or because insertion of candidate item $x$ into the shadow device really did lead to no increase in cardinality. This indicates that  ``good'' items would be rejected unnecessarily. This could be addressed by using multiple resets of the shadow HLL sketch, filling it with random items after each reset, and performing multiple insertions per candidate item $x$ to try to estimate the true value of $M[j]$ for each item $x$ (since after at least some of the resets followed by random insertions, we could expect $M[j]$ to increase when $x$ is inserted, and if $M[j]$ is small enough relative to the cardinality estimate, the change could be detected). The details of such an attack require further analysis which we do not pursue here, since we consider S1 to be the least realistic attack scenario.

%Finally, we briefly consider scenario S1 with an initially non-empty target HLL sketch. Since the target sketch has an unknown status for the adversary, our previous strategy for S1 of keeping many buckets empty does not work. One can attempt to use the shadow device to identify items that would result in small entries $M[j]$ (for unknown indices $j$) since such items are unlikely to increase the cardinality estimate of the target device by much.  This requires first filling the shadow device with sufficiently many items to escape from the ``small range correction'' phase. Then one could enter random items $x$ into the shadow device and check for cardinality increases. However this runs into the rounding issue described in Section~\ref{sec:scenarios-discussion}. Still, one could retain those items which do lead to a detectable change in cardinality and reject the others (and also estimate the value of $M[j]$ for the unknown bucket $j$ from the magnitude of the change, retaining only those for which $M[j]$ is small). But here the lack of a cardinality change could result either from rounding issues or because insertion of $x$ into the shadow device really did lead to no increase in cardinality. This indicates that  ``good'' items would be rejected unnecessarily. This could be addressed by using multiple resets, filling with random items, and then multiple insertions per item to try to estimate the true value of $M[j]$ for each item. The details of such an attack require further analysis which we do not pursue here, since we consider S1 to be the least realistic attack scenario.

@@

In scenarios S2 and S4 with an initially non-empty HLL sketch, we assume as for our earlier analysis of the standard HLL algorithm that all buckets have already been hit (this will be the case with high probability as soon as $m \log m$ items have been honestly inserted). Hence $C_0 = 0$ and $\sigma$ is evaluated at 0 in (\ref{eqn:ertl}), where its value is 0. This means that the estimate $\hat{\lambda}$ is computed using a denominator of the form:
\[
\sum_{k=1}^{\ell}C_k 2^{-k} + m \tau(1-C_{\ell+1}/m)2^{-\ell}.
\]
Now we can proceed exactly as in Sections~\ref{sec:nonemptyHLLS2} and~~\ref{sec:nonemptyHLLS4}, inserting items $x$ whose hashes have rightmost $\ell$ bits beginning with a ``1'' bit (and with a more relaxed condition in S4). These insertions do not alter the values of the $C_k$, and hence $\hat{\lambda}$ is unaffected. In S2, we expect to insert half the items in ${\mathcal{X}}$ without increasing the cardinality estimate $\hat{\lambda}$ at all. In S4, the analysis is the same as in Section~\ref{sec:nonemptyHLLS4}.

@@

We first begin by explaining the high-level idea of our approach. We allow an adversary ${\mathcal{A}}$ to interact with an HLL via various oracles that capture its capabilities in each of the three scenarios S2, S3 and S4: it may have an \texttt{insert($\cdot$)} oracle allowing it to add items to the HLL sketch, a \texttt{read} oracle that returns the HLL cardinality estimate, and a \texttt{corrupt} that returns all the HLL bucket contents (but not any secret key $K$); in S4 it has all three oracles. In addition, in all 3 scenarios, ${\mathcal{A}}$ has access to an \texttt{init($\cdot$)} oracle by which it sets the HLL parameters $n$, $\ell$, $\alpha_n$ used to initialise the HLL sketch. We assume ${\mathcal{A}}$ always makes exactly one \texttt{init($\cdot$)} query and that it is its first oracle call.

@@

The four oracles that  ${\mathcal{A}}$ can interact with behave as follows in the real world:
\begin{itemize}
\item \texttt{init($\cdot$)}: this oracle receives as input the HLL-$F$ parameters ($n$, $\ell$, $\alpha_n$) and performs initialisation of the HLL-$F$ sketch, including generating the array of buckets $M[j]$ and initialising all the entries to $0$, and selecting a key $K \leftarrow_{\$} \{0,1\}^k$ for keyed function $F: \{0,1\}^k \times {{\mathcal{D}}} \rightarrow  \{0,1\}^{n+\ell}$. This oracle has no output. It is available in all of S2, S3, S4. We assume it is only called once as ${\mathcal{A}}$'s first oracle query.
\item \texttt{insert($\cdot$)}: on input $x \in {{\mathcal{D}}}$, the oracle implements HLL-$F$ by setting $y = F_K(x)$ and following the procedure in Figure~\ref{fig:hll} to insert $x$ into the HLL sketch (that is, it uses the first $m$ bits of $y$ to determine a bucket index $j$ and the next $\ell$ bits of $y$ to determine a bit-string $w$; then $\rho(w)$ is used to update the bucket at index $j$ via $M[j]:= \max(M[j],\rho(w))$). This oracle has no output. It is available in all of S2, S3, S4. Without loss of generality (since HLL and variants of it we consider here have a deterministic insertion procedure), we assume all queries $x$ made by ${\mathcal{A}}$ to \texttt{insert($\cdot$)} are distinct.
\item \texttt{read}: this oracle has no input and returns the HLL-$F$ cardinality estimate as per Figure~\ref{fig:hll}, i.e.\ it computes a raw cardinality estimate via $E := \alpha_mm^2\cdot(\sum_{j=1}^m2^{-M[j]})^{-1}$ and then performs corrections for small and large values. We allow multiple queries to this oracle and make it available to ${\mathcal{A}}$ in S3 and S4, but not S2.
\item \texttt{corrupt}: this oracle has no input and returns the vector $(M[j])_{j=1}^{m}$ of HLL bucket contents (but not any secret key $K$). This oracle is available only in S4. We allow the adversary to query \texttt{corrupt} as often as it wishes during its attack. Our description of S4 in Section~\ref{sec:attacks} limited the adversary to at most one such query; here we allow multiple queries because doing so does not increase the difficulty of achieving security.  
\end{itemize}

@@

An HLL can leak information about items $x$ that were previously inserted. Consider the following attack from~\cite{cardestprivacy}: to test whether an item $x$ was previously inserted into the HLL by an honest user (say), the adversary makes the sequence of queries: \texttt{read}, \texttt{insert($x$)}, \texttt{read}. If the two cardinality estimates are different, the adversary can conclude definitively that $x$ was \emph{not} previously inserted into the HLL; if they are the same, it can estimate (with some error) that $x$ was previously inserted.\footnote{Errors can arise in two ways when $x$ was not previously inserted: either inserting $x$ changes a bucket value, but the change is masked by rounding when the cardinality estimate is computed, or inserting $x$ does lead to a change in a bucket value because $\rho(w)$ computed from $x$ is too small.} This attack still works even if HLL is implemented using a PRF and is unavoidable in view of the oracle queries (and in real settings, the API) available to the adversary. So such privacy leaks are an inherent feature of HLL if a rich enough API is provided. Note that this attack does not contradict our main result, which concerns prevention of cardinality manipulation rather than confidentiality.

%Still, simulation-based security notions are often used to model confidentiality properties, so it is tempting to think that our security definition might say something about confidentiality of the HLL inputs $x$. Our existing model does not, since ${\mathcal{A}}$ chooses all the inputs $x$ and so ``knows'' them already. However, it is easily modified to provide a confidentiality notion. For example, we can consider the insertion of honestly generated items $x$ that are drawn from some probability distribution $p$ on $\mathcal{D}$ in a first phase, with ${\mathcal{A}}$ then having access to only \texttt{read} and \texttt{corrupt} oracles in a second phase (we have to remove the \texttt{insert($\cdot$)} oracle in view of the general attack from~\cite{cardestprivacy}).

@@

An S4 adversary $\mathcal{A}$ can perform a series of pairs of \texttt{insert($x^i$)} and \texttt{corrupt} queries on chosen items $x^0, x^1, \ldots, x^{t-1}$. With high probability (assuming $t$ is not too large), each \texttt{insert} query hits a fresh bucket in the HLL sketch. Then comparing the results of the \texttt{corrupt} queries either side of the \texttt{insert($x^i$)} query reveals to $\mathcal{A}$ which bucket was hit by that  \texttt{insert} query. In turn, this leaks partial information about the corresponding UHF output $p_{x^i}(K)$, specifically, its $n$ least significant bits (since those bits are used to generate the bucket index for insertion). We now show how to exploit this partial information on $p_{x^i}(K)$ to recover $K$.

@@

%\todo{Sim based of indepednent interest, arguably more powerful than game-based notions with fixed winning predicates.}
%It is also an open problem to close the performance gap that exists between fast but non-cryptographic hashes used in HLL implementations today and the slower but cryptographically strong VIL-PRF that is needed to obtain security.
